\pagebreak

\section*{Sažetak}

Ovaj rad se bavi paralelizacijom algoritama sortiranja s ciljem optimizacije performansi kroz paralelno računarstvo. U eri velikih podataka, efikasno sortiranje velikih skupova podataka postaje ključno za aplikacije poput baza podataka i mašinskog učenja. Tradicionalni sekvencijalni algoritmi postaju usko grlo za velike skupove podataka.

Rad istražuje implementacije na višejezgrenim CPU-ima koristeći OpenMP i na GPU-ima koristeći CUDA. Implementirani su algoritmi: bitonic sort, merge sort, quick sort, radix sort i std::sort u sekvencijalnim i paralelnim varijantama.

Rezultati pokazuju značajna ubrzanja, posebno za velike skupove podataka, sa skalabilnošću na više jezgara. Poređenje CPU i GPU implementacija ističe prednosti GPU-a za paralelne algoritme.

Rad doprinosi razumijevanju primjene paralelnih tehnika na algoritme sortiranja i pruža uvide u optimizaciju performansi.

\vspace{1cm}
\textbf{Ključne riječi}: paralelizacija, algoritmi sortiranja, paralelno računarstvo, \\
CPU, GPU, OpenMP, CUDA

\section*{Abstract}

This thesis addresses the parallelization of sorting algorithms for performance optimization through parallel computing. In the era of big data, efficient sorting of large datasets is crucial for applications such as databases and machine learning. Traditional sequential algorithms become bottlenecks for large datasets.

The work explores implementations on multi-core CPUs using OpenMP and on GPUs using CUDA. Algorithms implemented include: bitonic sort, merge sort, quick sort, radix sort, and std::sort in sequential and parallel variants.

Results show significant speedups, especially for large datasets, with scalability across cores. Comparison of CPU and GPU implementations highlights GPU advantages for parallel algorithms.

The thesis contributes to understanding the application of parallel techniques to sorting algorithms and provides insights into performance optimization.

\vspace{1cm}
\textbf{Keywords}: parallelization, sorting algorithms, parallel computing, CPU, GPU, OpenMP, CUDA